---
title: "Guide to solveSimplex game theory solver"
author: "Carla M. Sette"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<p>&nbsp;</p>

## Overview
**[solveSimplex](https://github.com/cmsette/solveSimplex)** is an R function that describes the equilibrium dynamics of _single-population_ evolutionary game theory games of **2-5 strategies**. All strategies are assumed to exist in _admixture_, and engage in _continuous time replicator dynamics_. Methods were developed following Friedman and Sinervo (2016).

This code solves for any internal and face equilibria present in the system and classifies the equilibrium types (**source, saddle, sink**).

<p>&nbsp;</p>

## _Rock-Paper-Scissors_ as an example

### What is a population?
In a single-population game, members of the same group utilize different strategies to compete for the same prize. These competing strategies all interact with one another (exist in admixture), rather than distribute themselves into separate patches.

Examples of single-population games can be found in nature in every taxonomic group. Some examples include: 

- strains of _E. coli_ compete to have the largest colony
- male lizard morphs compete for mates (_Uta stansburiana_)
- salamander species compete to evade predators (_Taricha torosa_, _Ensatina eschscholzii xanthoptica_, _Ensatina eschscholzii eschscholzii_)

<p>&nbsp;</p>
### Rock-Paper-Scissors and the W-matrix
A **payoff matrix** is used to describe the competitive outcomes in a game. A payoff matrix has one row and one column for each strategy in the population. The **payoffs** are numerical values describing the fitness outcome to each strategy (by row) when in pairwise competition with every possible strategy (by column).

In the classic schoolyard game _Rock-Paper-Scissors_, players can choose from three strategies. For example, Rock receives a payoff of **0** (a tie) when in competition with other Rocks, a payoff of **-1** (defeat) against Paper, and a payoff of **1** (victory) against Scissors. 

Therefore, the payoff matrix for a game of Rock-Paper-Scissors looks like:
```{r echo=FALSE}
W <- matrix(c(0,-1,1, 1,0,-1, -1,1,0), ncol=3, byrow=T)
colnames(W) <- rownames(W) <- c("Rock", "Paper", "Scissors")
print(W)
```

<p>&nbsp;</p>
### The replicator equation
In a population subject to selection, the proportion of the population belonging to each strategy may change over time. If a strategy experiences a higher average fitness than the population average, it will increase in frequency; if a strategy experiences a lower average fitness, it will decrease.

Under continuous time replicator dynamics (i.e. population frequencies that change continuously over time), the average fitness ($w_{i}$) of a strategy (i) in competition with all other strategies (j) can be calculated as the weighted average of the payoff ($w_{ij}$) against each strategy, based on the relative frequency of each strategy in the population ($s_{j}$): 

                            1)    $w_{i}(s)=∑\begin{array}{c} _n \\ ^{j=1}\end{array} ( w_{ij} s_{j} )$

The average population fitness is each strategy's average fitness ($w_{i}$), weighted by its frequency in the population ($s_{i}$):

                            2)    $\bar{w}=∑w_{i} s_{i}$

Finally, the replicator equations describe how the frequency of each strategy in the population will increase or decrease over time $\dot{s}_{i}(t)$, based on the difference between the fitness of that strategy ($w_{i}$) and the average fitness of all strategies in the population ($\bar{w}$).

                            3)    $\dot{s}_{i}(t)=(w_{i}-\bar{w})s_{i}(t)$

At equilibrium $s^*$, all strategies have equal average fitness. A system of 3+ strategies may have an internal equilibrium, as well as equilibria on the faces (where at least 1 strategy has a population of 0). Conversely, not every system of strategies has an internal equilibrium at all. 

<p>&nbsp;</p>
### Classifying equilibria
The existence of an equilibrium does not guarantee a stable population. By classifying equilibria (when they exist), we can predict how the frequencies of each strategy will change in a population over time. This allows us to make predictions from evolutionary game theory models!

To classify an equilibrium point, we describe the trajectories around the equilibrium point. An n x n Jacobian matrix is calculated from the set of replicator equations $\dot{s}_{i}(t)$ at equilibrium ($s_{j}$): 

                            4)    $Dw_i(s^* ) = ∂w_i(s^* )/∂s_j$

We scale the Jacobian matrix by multiplying it by a projection matrix: 

                            5)    $P_0 = I-(\frac{1}{n} * 1_{nxn})$

Finally, the eigenvalues for this matrix are calculated and sorted by their real parts from largest to smallest (_sensu_ Sandholm 2010). 

                            6)    $J = Dw_i (s^* ) \cdot P_0$

There should be one eigenvalue equal to 0, and n-1 nonzero eigenvalues. Positive eigenvalues represent trajectories moving away from an equilibrium point, while negative eigenvalues represent trajectories moving toward an equilibrium point. 

- **solveSimplex** returns **No eq** when there is no equilibrium present at the level of the game (internal equilibrium or face) being assessed. 
- **Sink** equilibria are attractors - populations with any starting frequency of strategies will move toward the equilibrium frequency. The nonzero eigenvalues will all have negative real parts. 
- **Source** equilibria are repellers - if perturbed away from the equilibrium frequency, the population will not return to equilibrium. The nonzero eigenvalues will all have positive real parts. 
- **Saddle** equilibria act as attractors from some trajectories and repellers from others. The eigenvalues will have both positive and negative real parts. **solveSimplex** classifies saddles by the number of positive eigenvalues present (eg. **Saddle1, Saddle2, Saddle3**).
- Finally, this method of classifying games is **Inconclusive** when all the eigenvalues are zero. This occurs in the presence of a _strange attractor_ or population trajectories that _orbit_ around a Nash equilibrium (sink).

<p>&nbsp;</p>
### Solving Rock-Paper-Scissors
Because the payoffs for each strategy in **Rock-Paper-Scissors** are balanced, each strategy has equal fitness and the equilibrium is $s^*$ = {$\frac{1}{3}$ Rock, $\frac{1}{3}$ Paper, $\frac{1}{3}$ Scissors}. The internal equilibrium is classified as **inconclusive**. The reason for this is made clear when we look at the 2-strategy faces of the game.
```{r echo=FALSE}
Equilibria <- data.frame("R" = c(0.3333333,0,1,0), "P" = c(0.3333333,1,0,0), "S" = c(0.3333333,0,0,1), "Eq" = c("Inconclusive","Saddle1","Saddle1","Saddle1"), "Int" = c("yes",rep("",3)), "Apo" = rep("",4), "A.Apo" = rep("",4))
rownames(Equilibria) = c("R.P.S", "R.P.0", "R.0.S", "0.P.S")
Equilibria <- list("Equilibria"=Equilibria)
Equilibria
```

In competition between pairs of strategies, there is always a winner and a loser (eg. Paper beats Rock). The pairs of 2 strategies make up the faces of the game. When each pair is assessed on its own, the losing strategy is a source and the winning strategy is a **sink**. 
```{r echo=FALSE}
faces <- data.frame("R" = c(0,1,NA), "P" = c(1,NA,0), "S" = c(NA,0,1), "Eq" = c("Sink","Sink","Sink"))
rownames(faces) = c("R.P._", "R._.S", "_.P.S")
faces <- list("Faces(2)"=faces)
faces
```

When the 2-strategy faces are assessed as a part of the full game, each of the vertices becomes a **saddle** due to the trajectory leading to this point along one 2-strategy face and the trajectory leading away from this point along the next 2-strategy face. (Paper beats Rock, but Scissors in turn beats Paper.)

Similar to the trajectories along the faces of this game, population frequencies will orbit around the internal equilibrium, therefore the eignvalue method of classification is **inconclusive**.

<p>&nbsp;</p>
### Other game features
The solveSimplex function also returns whether a game (and in some cases its faces) is **intransitive** and also whether it has **apostasis** (negative frequency-dependence) or **anti-apostasis** (positive frequency-dependence).

With strict-sense **intransitivity**, the competitive hierarchy in the system is cyclical. Each strategy is the best response (clear winner) to the same number of other strategies. Competitive intransitivity is often a feature of stable systems, those with sink equilibria.

**Apostasis** and **Anti-apostasis** are terms from biology that describe negative frequency-dependence (also known as rare advantage) and positive frequency-dependence. Rare advantage contributes to system stability as the fitness of apostatic strategies drops as these strategies become more abundant.

<p>&nbsp;</p>
## Using the [solveSimplex](https://github.com/cmsette/solveSimplex) function
```{r echo=F}
print("solveSimplex(W, names=c(\"R\", \"P\", \"S\"), deconstruct=F, debug=F)", quote=F)
```

<p>&nbsp;</p>
### The W matrix
The W matrix (**W**) can either be a matrix of payoff values or a vector of payoff valuess arranged by row.
```{r echo=F}
print("W <- matrix(c(0,-1,1, 1,0,-1, -1,1,0), ncol=3, byrow=T)", quote=F)
```
```{r echo=FALSE}
W <- matrix(c(0,-1,1, 1,0,-1, -1,1,0), ncol=3, byrow=T)
print(W)
```
Or just:
```{r echo=F}
print("W <- c(0,-1,1, 1,0,-1, -1,1,0)", quote=F)
```

<p>&nbsp;</p>
### Names
You can enter a vector of custom names for the strategies in your system. Otherwise the default initials will be used (**R**ock-**P**aper-**S**cissors-**L**izard-Spoc**K**).
```{r echo=F}
print("names=c(\"R\", \"P\", \"S\")", quote=F)
```

<p>&nbsp;</p>
### Solving face games
In some cases, you may be interested in whether more complex systems of competitors will find a new stable equilibrium if one or more competitors is removed (hysteresis). Likewise, you may be interested whether a system of competitors was able to evolve from a smaller system. 

The solveSimplex function will additionally assess all face games as independent games when the deconstruct parameter is set to **TRUE** (default). This can be suppressed by changing this to **FALSE**.
```{r echo=F}
print("deconstruct=F", quote=F)
```

<p>&nbsp;</p>
### Debugging
The Jacobian matrix eigendecomposition of every game assessed (full and face games, as well as independent face games) will be returned when the debug parameter is set to **TRUE**.
```{r echo=F}
print("debug=T", quote=F)
```